---
title: "Prediction of the presence of *Aedes aegypti* in the Metropolitan Area of Mexico City with LightGBM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prediction of the presence of Aedes aegypti in the Metropolitan Area of Mexico City with LightGBM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Objective

Predicting the probability of the presence of *Ae aegypti* in Metropolitan Area of Mexico City using LightGBM

## Step 1. Load the pkg

```{r setup}
Sys.setenv("_R_CHECK_LIMIT_CORES_" = "false")
library(tidyverse) 
library(magrittr) 
library(tidymodels) # Machine Learning en R
library(skimr) 
library(bonsai)
library(themis) # Upsampling - downsampling
library(parallel)
library(doParallel)
library(sf)
```

## Step 2. Load the dataset and the area of interes

```{r load_dataset, warning=FALSE, message=FALSE}
# Load the aedes aegypti dataset ####

data <- aeaegypticdmx::ae_aegypti_cdmx


# Step. extract the area of interes ####
aoi <-  aeaegypticdmx::aoi

```

### Step 2.1. visualization of data and aoi

```{r map_dengue_cases}
#| warning: False
#| message: False
#| echo: true
#| output: true
#| out.width: "100%"
library(mapgl)
maplibre(style = carto_style("positron"),
         bounds = data) |>
    add_circle_layer(id = "Aedes aegypti",
                     source = data,
                     circle_color = match_expr(
                         column = "class",
                         values = c("presence", "pseudoabs"), # reemplaza por tus valores
                         stops = c("#E01A59", "#0F9D58")),
                     circle_stroke_color = "white", # color = "white"
                     circle_stroke_width = 0.5, # lwd = 0.5
                     circle_radius = 3) |>
    #    Capa 2 — polígono aoi (sin leyenda)
    add_fill_layer(id = "aoi",
                   source = aoi,
                   fill_color = "#ECB32D",
                   fill_opacity = 0.5, # ajusta al gusto
                   fill_outline_color = "black") |>
    add_line_layer(id           = "aoi-borde",
                   source       = aoi,      # misma fuente
                   line_color   = "black",  # color = "black"
                   line_width   = 2,        # lwd = 2
                   line_opacity = 1) |>
    add_categorical_legend(legend_title = "Aedes aegypti",
                           position = "bottom-left",
                           values = c("presence", "pseudoabs"), 
                           colors = c("#E01A59", "#0F9D58")) |>
    add_fullscreen_control(position = "top-left") |> 
    add_navigation_control()
```

### Step 2.2. select the data of year 2024

```{r select_data,  warning=FALSE, message=FALSE}
data <- data 
```

### Step 2.3. Glimpse of your data

```{r glimpse,  warning=FALSE, message=FALSE}
data |>
    dplyr::glimpse()
```

### Step 2.3. EDA Univariate

```{r skim,  warning=FALSE, message=FALSE}
# EDA Univariado ####
skimr::skim(data |> sf::st_drop_geometry())
```

```{r gt_plt_summary,  warning=FALSE, message=FALSE}
gtExtras::gt_plt_summary(data |> sf::st_drop_geometry())
```

### Step 3.4. EDA Multivariate

### 3.5. Weights

#### table

```{r balanceo_table, warning=FALSE, message=FALSE }
# Balanceo
data |>
    sf::st_drop_geometry() |>
    dplyr::group_by(class) |> 
    dplyr::count(name = 'frec') |>
    dplyr::ungroup() |>
    dplyr::mutate( Porc= frec/sum(frec)) |>
    gt::gt()

```

#### graph

```{r graph,warning=FALSE, message=FALSE}
data |>
    sf::st_drop_geometry() |>
    dplyr::group_by(class) |> 
    dplyr::count(name = 'frec') |>
    dplyr::ungroup() |>
    dplyr::mutate(Porc= frec/sum(frec)) |>
    ggplot2::ggplot(ggplot2::aes(x= class, 
                                 y= Porc)) +
    ggplot2::geom_segment(ggplot2::aes(xend = class, 
                                       y = 0, 
                                       yend=Porc), 
                          color= c("#4285F4","#E01A59"), 
                          linewidth= 1) +
    ggplot2::geom_point(size=5, 
                        color= c("#4285F4","#E01A59")) +
    ggplot2::coord_flip() +
    ggplot2::scale_y_continuous(labels = scales::percent_format()) +
    ggplot2::labs(title= 'Porcentaje de Registros', 
                  y = "Porcentaje", 
                  x = "") +
    ggplot2::theme_bw()
```

## Step 4. Modeling

### Step 4.1. Step Train-Test Split

#### initial split

```{r train_test, warning=FALSE, message=FALSE}
# . Train-Test Split ####
set.seed(12345) # Semilla para aleatorios
split <- data |>
    rsample::initial_split(prop = 0.8, 
                           strata = class)
```

#### Train

```{r train}
train <- rsample::training(split)
dim(train)
```

#### Test

```{r test}
test <- rsample::testing(split)
dim(test)
```

### Step 4.2. Preprocess

#### Importance Weights and recipe

```{r preprocess_recipe, warning=FALSE, message=FALSE}
receta <- train |>
    sf::st_drop_geometry() |>
    #dplyr::select(class, dplyr::starts_with("bio")) |>
    recipes::recipe(class ~ . ) |>## Crea la receta 
    ## Eliminar variables que no usaremos
    # step_rm() |>
    ## Crear nuevas variables (insight desde el EDA)
    #recipes::step_mutate(temperature = temperature/1000) |>
    ## Imputar los datos 
    # step_impute_mean()
    recipes::step_impute_knn(recipes::all_predictors() ) |>
    ## Estandarizacion/Normalizacion de numericas
    recipes::step_normalize(recipes::all_numeric(),
                            -recipes::all_outcomes()) |>
    ## Crear una categoría "otros" que agrupe a categorias pequeñas
    recipes::step_other(recipes::all_nominal(), 
                        -recipes::all_outcomes() , 
                        threshold = 0.07, 
                        other = "otros") |>
    ## Crear una categoría "new" para observaciones con labels "no muestreados"
    recipes::step_novel(recipes::all_nominal(), 
                        -recipes::all_outcomes(), 
                        new_level = "new") |>
    ## Crear variables indicadoras para cada categoría
    recipes::step_dummy(recipes::all_nominal(), 
                        -recipes::all_outcomes()) |># Dummy
    ## Eliminar automáticamente variables con alta correlacion 
    ## para evitar la multicolinealidad xi ~ xj
    recipes::step_corr(recipes::all_numeric(),
                       -recipes::all_outcomes(), 
                       threshold = 0.80) |>
  recipes::step_nzv(all_predictors()) |>
    # Tambien podemos eliminar variables con multicolinealidad "a mano"
    #recipes::step_rm(suhi_night, suhi_day) |>
    ## Balancear usando upsampling
    ## over_ratio implica que vamos a llevar a la clase minoritaria a 
    ## alrededor de 90% de filas que la mayoritaria
    themis::step_upsample(class, 
                          over_ratio= 0.99, 
                          skip= TRUE, 
                          seed = 345)
```

```{r recipe_view, warning=FALSE, message=FALSE}
receta

receta |>
    recipes::prep() |>
    recipes::bake(new_data = NULL) |>
    dplyr::glimpse()
```

### Step 4.3. Hyperparameter Optimization and Training

#### Resampling with with spatial block cross-validation

```{r remuestreo, warning=FALSE, message=FALSE}
set.seed(12345)
library(tidysdm)
sp_block_cv  <- TRUE
if(sp_block_cv == TRUE){
cv <- spatialsample::spatial_block_cv(train,
                                      method = "random",
                                      repeats = 2,
                                      v = 10)
autoplot(cv) +
    ggplot2::geom_sf(data = aoi,
                     fill = NA,
                     col = "black",
                     lwd = 0.5)
} else{
cv <- rsample::vfold_cv(train |> sf::st_drop_geometry(), 
                        v = 10, 
                        repeats = 2, 
                        strata = class)
}
cv
```

```{r}
if(sp_block_cv == TRUE){
check_splits_balance(cv, class)
} else{
}

```

#### Metrics

```{r metricas,  warning=FALSE, message=FALSE}
#. Métricas ####
library(yardstick)
library(tidysdm)
metricas <- yardstick::metric_set(roc_auc,
                                  yardstick::sens,
                                  yardstick::spec,
                                  yardstick::brier_class,
                                  yardstick::pr_auc,
                                  bal_accuracy,
                                  yardstick::pr_auc, 
                                  accuracy, 
                                  tidysdm::boyce_cont, 
                                  tidysdm::tss,
                                  tidysdm::tss_max,
                                  kap)
metricas
```

#### Activate Parallelization

```{r activate_parallelization, warning=FALSE, message=FALSE}
parallel::detectCores(logical=FALSE)

cl <- parallel::makePSOCKcluster(8)
doParallel::registerDoParallel(cl)
```

#### Model specification

```{r esp_mod, warning=FALSE, message=FALSE}
# =============================================================================
#  Model specification
# =============================================================================
library(bonsai)
lgbm_sp <- parsnip::boost_tree(mtry = tune::tune(), 
                               trees = tune::tune(), 
                               tree_depth = tune::tune(),
                               #min_n          = tune::tune(), 
                               #sample_size    = tune::tune(), 
                               loss_reduction = tune::tune(), 
                               learn_rate= tune::tune()) |>
    parsnip::set_engine("lightgbm") |>
    parsnip::set_mode("classification")
```

#### Workflow

```{r worflow,  warning=FALSE, message=FALSE}
# =============================================================================
# Workflow
# =============================================================================
lgbm_wflow <- workflows::workflow() |>
    workflows::add_recipe(receta) |>
    workflows::add_model(lgbm_sp) 
lgbm_wflow
```

#### Hyperparameter Space

##### Search Mesh

```{r mesh_search, warning=FALSE, message=FALSE}
# =============================================================================
# Hyperparameter Space
# =============================================================================
set.seed(12345)
lgbm_grid <- lgbm_sp |>
  workflowsets::extract_parameter_set_dials() |>
  recipes::update(mtry       = dials::mtry(range= c(3, 7)),
                  trees      = dials::trees(range = c(100, 600)),
                  tree_depth = dials::tree_depth(range= c(3, 7))) |>
  dials::grid_space_filling(size = 150, 
                            type = "latin_hypercube")

lgbm_grid 
```

#### Training

```{r entrenamiento_mesh_search, warning=FALSE, message=FALSE}
# =============================================================================
# training
# =============================================================================
tictoc::tic()
set.seed(12345)
lgbm_tuned <- tune::tune_grid(lgbm_wflow, ## Model
                             resamples= cv, ## Crossvalidation
                             grid = lgbm_grid, ## grid search
                             metrics = metricas, ## Metrics
                             control= tune::control_grid(allow_par = T, 
                                                         verbose  = T,
                                                         save_pred = T))
tictoc::toc()

```

Sensitivity

```{r show_best_sens}
tune::show_best(x = lgbm_tuned, metric = 'sens', n = 10)
```

Accuracy

```{r show_best_accuracy}
tune::show_best(lgbm_tuned, metric = 'accuracy', n = 10)
```

Specificity

```{r show_best_spec}
tune::show_best(lgbm_tuned, metric = 'spec', n = 10)
```

### Step 4.4. Final Model

#### Define the best combination of hyperparameters

#### In this workflow, the best combination of hyperparameters was defined using sensitivity, because the goal is to predict where dengue cases are located.

```{r best_combination, warning=FALSE, message=FALSE}
lgbm_pars_fin <- tune::select_best(lgbm_tuned, 
                                   metric = "sens")
lgbm_pars_fin
```

#### Finalize (give values to tunable parameters) the workflow

```{r finalize_workflow, warning=FALSE, message=FALSE}
lgbm_wflow_fin <- lgbm_wflow |>
    tune::finalize_workflow(lgbm_pars_fin)
lgbm_wflow_fin
```

### Step 4.5. Train the final model

```{r train_final_model,warning=FALSE, message=FALSE}
lgbm_fitted <- fit(lgbm_wflow_fin, train)
lgbm_fitted
```

## Step 5. Internal Validation

### Step 5.1 Metrics

```{r evaluate_final_model,warning=FALSE, message=FALSE}
denriskmxnonendemic::class_metrics(ml = lgbm_fitted,
                                   train = train,
                                   test = test,
                                   ml_title = "LightGBM Aedes aegypti (2015-2025)")
```

### Step 5.2 Correlation Tetrácorica

#### Train

```{r cor_tetrachoric_train, warning=FALSE, message=FALSE}
y  <- train %>%
    predict(lgbm_fitted , new_data = ., type = "prob") |>
    dplyr::mutate(Real= train$class) |>
    dplyr::mutate(Real = ifelse(Real == "presence", 1, 0)) |>
    dplyr::mutate(pred_presence = ifelse(.pred_presence >= .5, 1, 0)) |>
    dplyr::select(-.pred_pseudoabs)
correlation::correlation(data = y,
                         select = "Real",
                         select2 = "pred_presence",
                         method = "tetrachoric")
```

#### Test

```{r cor_tetrachoric_test, warning=FALSE, message=FALSE}
x  <- test %>%
    predict(lgbm_fitted , new_data = ., type = "prob") |>
    dplyr::mutate(Real= test$class) |>
    dplyr::mutate(Real = ifelse(Real == "presence", 1, 0)) |>
    dplyr::mutate(pred_presence = ifelse(.pred_presence >= .5, 1, 0)) |>
    dplyr::select(-.pred_pseudoabs)
correlation::correlation(data = x,
                         select = "Real",
                         select2 = "pred_presence",
                         method = "tetrachoric")
```


## Step 7. Post-analysis

```{r vip, warning=FALSE, message=FALSE}
lgbm_fitted |>
workflows::extract_fit_parsnip() |>
    vip::vip(geom= "col",
             num_features = 16,
              aesthetics = list(color = "white",
                                fill = "black",
                                size = 1))
```

## Step 8. Prediction

### Step 8.1. Load the layer

```{r lyrs, warning=TRUE, message=FALSE}
library(tidyterra)
library(terra)

lyrs <- terra::rast(system.file("extdata",
                                  "raster_layers_cdmx.tif",
                                  package = "aeaegypticdmx"))  |>
    dplyr::select(-dplyr::contains("2023")) |>
    dplyr::rename(hfp = hfp2022,
                  pop = pop_2024,
                  indexp = indexp2022,
                  temp = temperature_2024,
                  built = built_2024,
                  tree = tree_2024,
                  ndvi = ndvi_2024,
                  psdi = psdi_2024,
                  prcp = prcp_2024,
                  tmin = tmin_2024,
                  tmax = tmax_2024,
                  eddi = eddi_2024,
                  hr = hr_2024,
                  suhi = suhi_2024,
                  uhi = uhi_2024)

```

### Step 8.2. Prediction

```{r pred, warning=FALSE, message=FALSE}
# Make the prediction
pred <- tidysdm::predict_raster(object = lgbm_fitted, 
                                raster = lyrs,
                                type = "prob")
# Extract the predictions from the area of interest
pred <- terra::crop(x = pred,
                    y = aoi,
                    mask = TRUE)
```

### Step 8.3. Visualization

```{r pred_visualization, warning=FALSE, message=FALSE}
library(mapgl)
library(terra)
maplibre_view(pred[[1]], 
              palette = viridis::turbo) |>
    add_continuous_legend(legend_title = "Probabilidad",
                          position = "bottom-left",
                          values = c(0, 0.25, 0.50, 0.75, 1),
                          colors = viridis::turbo(n = 10)) |>
  add_fullscreen_control(position = "top-left") |> 
  add_navigation_control()

```

### Step 9. Stop Cluster

```{r stop_cl}
parallel::stopCluster(cl)
```

### Step 10. Session Info

```{r session_info}
sessioninfo::session_info()
```







